---
title: "Somatic Variants – Clinical Report"
date: "`r format(Sys.time(), '%d %B %Y')`"
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    css: clinical-report.css
    include-after-body: clinical-report-extra.html
    theme: flatly
params:
  somatic_rds:   "somatic_mutations.rds"
  qc_pdf:        "variants_on_territories.pdf"
  tmb_csv:       "TMB.csv"
  maf_pdf:       "max_MAF_per_territories.pdf"
  clinical_csv:  "clinical_reports.csv"
  multiqc_html:  "multiqc_report.html"
  base_path:     "."
  example_mode:  true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(ggplot2)
  library(knitr)
  library(kableExtra)
})

# ── Resolve file paths ────────────────────────────────────────────────────────
bp <- params$base_path

resolve <- function(p) {
  if (is.null(p) || p == "" || p == "NO_FILE") return(NULL)
  if (file.exists(p)) return(p)
  joined <- file.path(bp, basename(p))
  if (file.exists(joined)) return(joined)
  NULL
}

f_rds         <- resolve(params$somatic_rds)
f_tmb         <- resolve(params$tmb_csv)
f_clinical    <- resolve(params$clinical_csv)
f_multiqc     <- resolve(params$multiqc_html)

example_mode  <- params$example_mode ||
  (is.null(f_rds) && is.null(f_tmb))

# ── Helper: coloured badge ────────────────────────────────────────────────────
badge <- function(label, type = "pass") {
  cls <- switch(type, pass = "badge-pass", warn = "badge-warn", fail = "badge-fail", "badge-pass")
  htmltools::HTML(sprintf('<span class="badge %s">%s</span>', cls, label))
}
```

::: {.report-header}
# Somatic Variants Clinical Report

::: {.subtitle}
Somatic Variants Pipeline &nbsp;|&nbsp; `r format(Sys.time(), "%d %B %Y")`
:::
:::

```{r meta-banner, results='asis'}
if (example_mode) {
  cat('<div class="summary-box warning">',
      '<strong>⚠ Example / Demo Mode</strong> – Pipeline outputs were not found.',
      'This report shows illustrative mock data for layout verification.',
      '</div>')
}
```

---

## 1. Clinical Data

```{r clinical-table}
if (!is.null(f_clinical) && file.exists(f_clinical)) {
  clin <- tryCatch(read_csv(f_clinical, show_col_types = FALSE), error = function(e) NULL)
} else {
  clin <- NULL
}

if (is.null(clin)) {
  clin <- data.frame(
    Parameter = c("Patient ID", "Sample ID", "Kit / Panel", "Project",
                  "Report date", "Pipeline version"),
    Value     = c("PT-EXAMPLE", "S-001", "Whole Exome", "Demo project",
                  format(Sys.Date(), "%d %b %Y"), "somatic_variants_pipeline v1.0")
  )
}

kable(clin, col.names = c("Parameter", "Value"),
      format = "html", escape = FALSE) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

---

## 2. Executive Summary

```{r exec-summary}
n_somatic <- NA_integer_
tmb_val   <- NA_real_

if (!is.null(f_rds) && file.exists(f_rds)) {
  tryCatch({
    dat <- readRDS(f_rds)
    n_somatic <- nrow(dat)
  }, error = function(e) NULL)
}

if (!is.null(f_tmb) && file.exists(f_tmb)) {
  tryCatch({
    tmb_tbl <- read_csv(f_tmb, show_col_types = FALSE)
    if ("TMB" %in% colnames(tmb_tbl)) tmb_val <- round(tmb_tbl$TMB[1], 2)
  }, error = function(e) NULL)
}

if (is.na(n_somatic)) n_somatic <- 47L   # example
if (is.na(tmb_val))   tmb_val   <- 4.2   # example

tmb_class <- if (tmb_val >= 10) "High (≥10 mut/Mb)" else if (tmb_val >= 3) "Intermediate (3–10 mut/Mb)" else "Low (<3 mut/Mb)"
tmb_badge <- if (tmb_val >= 10) badge("High", "fail") else if (tmb_val >= 3) badge("Intermediate", "warn") else badge("Low", "pass")
```

::: {.summary-box}
**Total somatic variants detected:** `r n_somatic`  
**Tumour mutational burden (TMB):** `r tmb_val` mut/Mb — `r tmb_class` `r tmb_badge`
:::

---

## 3. MultiQC Quality Control {#multiqc-qc}

```{r multiqc-section, results='asis'}
if (!is.null(f_multiqc) && file.exists(f_multiqc)) {
  cat(sprintf(
    '<div class="summary-box pass"><strong>✔ MultiQC report available.</strong>
    <a href="%s" target="_blank">Open full interactive MultiQC report</a></div>',
    basename(f_multiqc)
  ))
} else {
  cat('<div class="summary-box warning">MultiQC report not found.
  Run the pipeline with <code>skip_multiqc = false</code> to generate QC aggregation.</div>')
}
```

### 3.1 Pre-Alignment QC (FastQC)

```{r fastqc-plots, results='asis'}
plot_dir <- file.path(bp, "docs/test_report/plots")
fq_qual  <- file.path(plot_dir, "fastqc_per_base_quality.png")
fq_adapt <- file.path(plot_dir, "fastqc_adapter_content.png")

show_img <- function(path, alt = "") {
  if (file.exists(path)) {
    cat(sprintf('<div class="plot-container"><img src="%s" alt="%s"/></div>', path, alt))
  }
}

if (file.exists(fq_qual) || file.exists(fq_adapt)) {
  show_img(fq_qual,  "Per-base sequence quality")
  show_img(fq_adapt, "Adapter content")
} else {
  cat("*FastQC plots not found. Run the pipeline to generate.*")
}
```

### 3.2 Post-Alignment QC (samtools stats, duplicates)

```{r aln-plots, results='asis'}
aln_plot  <- file.path(plot_dir, "alignment_stats.png")
dup_plot  <- file.path(plot_dir, "duplication_rate.png")
ins_plot  <- file.path(plot_dir, "insert_size_distribution.png")

if (any(file.exists(c(aln_plot, dup_plot, ins_plot)))) {
  show_img(aln_plot, "Alignment summary")
  show_img(ins_plot, "Insert size distribution")
  show_img(dup_plot, "Duplication rate")
} else {
  cat("*Post-alignment QC plots not found. Run the pipeline to generate.*")
}
```

### 3.3 Post-Variant-Calling QC (bcftools stats)

```{r vc-plots, results='asis'}
vc_plot <- file.path(plot_dir, "variant_calling_stats.png")
if (file.exists(vc_plot)) {
  show_img(vc_plot, "Variant calling stats")
} else {
  cat("*Variant calling QC plots not found. Run the pipeline to generate.*")
}
```

---

## 4. Sequencing Results

```{r seq-results}
if (!is.null(f_rds) && file.exists(f_rds)) {
  tryCatch({
    dat <- readRDS(f_rds)
  }, error = function(e) { dat <<- NULL })
} else {
  dat <- NULL
}

if (is.null(dat)) {
  dat <- data.frame(
    sample_id       = "S-001",
    Chr             = c("chr7", "chr17", "chr3", "chr12"),
    Start           = c(55174773L, 7674220L, 178936091L, 25398284L),
    End             = c(55174773L, 7674220L, 178936091L, 25398284L),
    Ref             = c("A", "C", "C", "G"),
    Alt             = c("T", "T", "T", "A"),
    Gene.refGene    = c("EGFR", "TP53", "PIK3CA", "KRAS"),
    ExonicFunc.refGene = c("nonsynonymous SNV", "nonsynonymous SNV",
                           "nonsynonymous SNV", "nonsynonymous SNV"),
    AAChange.refGene = c("p.L858R", "p.R248W", "p.E545K", "p.G12D"),
    FREQ            = c(0.42, 0.35, 0.18, 0.55),
    DP              = c(312L, 198L, 245L, 420L),
    method          = c("MuTect,VarScan", "MuTect", "VarScan", "MuTect,VarScan"),
    stringsAsFactors = FALSE
  )
}
```

### 4.1 Somatic Summary

```{r somatic-summary}
if (nrow(dat) > 0) {
  summ <- dat %>%
    summarise(
      `Total variants`     = n(),
      `Coding SNVs`        = sum(grepl("nonsynonymous|stop|start", ExonicFunc.refGene, ignore.case = TRUE), na.rm = TRUE),
      `Indels`             = sum(grepl("insert|delet|frameshift", ExonicFunc.refGene, ignore.case = TRUE), na.rm = TRUE),
      `Median VAF (%)`     = round(median(FREQ, na.rm = TRUE) * 100, 1),
      `Median depth`       = round(median(DP, na.rm = TRUE), 0),
      .groups = "drop"
    )
  kable(summ, format = "html") %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
}
```

### 4.2 Variant Table

```{r variant-table}
display_cols <- intersect(
  c("sample_id", "Gene.refGene", "Chr", "Start", "Ref", "Alt",
    "ExonicFunc.refGene", "AAChange.refGene", "FREQ", "DP", "method"),
  colnames(dat)
)

dat_disp <- dat[, display_cols, drop = FALSE]
colnames(dat_disp) <- gsub("\\.refGene$", "", colnames(dat_disp))

if ("FREQ" %in% colnames(dat_disp)) {
  dat_disp$FREQ <- paste0(round(dat_disp$FREQ * 100, 1), "%")
}

kable(dat_disp, format = "html", escape = FALSE) %>%
  kable_styling(
    full_width        = TRUE,
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  ) %>%
  scroll_box(width = "100%", height = "400px")
```

---

## 5. TMB Analysis

```{r tmb-plot, fig.width=7, fig.height=4}
tmb_df <- tryCatch({
  if (!is.null(f_tmb) && file.exists(f_tmb)) read_csv(f_tmb, show_col_types = FALSE)
  else NULL
}, error = function(e) NULL)

if (is.null(tmb_df)) {
  tmb_df <- data.frame(sample_id = "S-001", TMB = 4.2)
}

ref_lines <- c(1, 10, 100)

ggplot(tmb_df, aes(x = factor(sample_id), y = TMB)) +
  geom_bar(stat = "identity", fill = "#2e86c1", width = 0.5) +
  geom_hline(yintercept = ref_lines, linetype = "dashed",
             color = c("#27ae60", "#e67e22", "#e74c3c"),
             linewidth = 0.6) +
  annotate("text", x = Inf, y = ref_lines + c(0.3, 0.6, 3),
           label = paste0(ref_lines, " mut/Mb"), hjust = 1.1,
           size = 3, color = c("#27ae60", "#e67e22", "#e74c3c")) +
  labs(title = "Tumour Mutational Burden (TMB)",
       x = "Sample", y = "TMB (mutations / Mb)") +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

## 6. Somatic Plots

```{r somatic-plots, results='asis'}
qc_pdf_path  <- resolve(params$qc_pdf)
maf_pdf_path <- resolve(params$maf_pdf)

for (pdf_path in c(qc_pdf_path, maf_pdf_path)) {
  if (!is.null(pdf_path) && file.exists(pdf_path)) {
    cat(sprintf(
      '<p><a href="%s" target="_blank">📄 Open %s</a></p>\n',
      pdf_path, basename(pdf_path)
    ))
  }
}

plot_dir2 <- file.path(bp, "docs/test_report/plots")
ins2 <- file.path(plot_dir2, "insert_size_distribution.png")
if (!is.null(qc_pdf_path) && !file.exists(qc_pdf_path) &&
    !file.exists(ins2)) {
  cat("*QC plots not available. Run the pipeline to generate.*")
}
```

---

## 7. Copy Number Variants (CNV)

```{r cnv-section, results='asis'}
cat('<div class="summary-box">CNV analysis is not yet integrated into this pipeline run.
    Please refer to dedicated CNV workflow outputs if available.</div>')
```

---

## 8. Structural Variants

```{r sv-section, results='asis'}
cat('<div class="summary-box">Structural variant calling is not performed in this run.</div>')
```

---

## 9. Germline Variants

```{r germline-section, results='asis'}
cat('<div class="summary-box">Germline variant analysis is not included in this somatic-only run.</div>')
```

---

## 10. Methods

The somatic variants analysis was performed using the **Somatic Variants Pipeline** (Nextflow DSL2).

**Upstream processing** (from FASTQ):

1. **Quality trimming** – Cutadapt (adapter trimming) + seqtk trimfq (base-quality trimming)
2. **Alignment** – BWA-MEM to hg38 reference
3. **BAM processing** – Picard AddOrReplaceReadGroups, Picard MarkDuplicates, GATK BaseRecalibrator + PrintReads (BQSR)
4. **Variant calling** – MuTect 1.x and VarScan 2 (mpileup2snp / mpileup2indel)
5. **Annotation** – ANNOVAR (refGene, dbSNP, COSMIC, ClinVar, CADD)
6. **QC aggregation** – FastQC (pre/post-trim), samtools stats, bcftools stats, MultiQC

**Downstream processing** (from pretty CSV):

1. **Target/bait annotation** – bedtools intersect with target and bait BED files
2. **Candidate selection** – frequency/depth filtering, cancer gene list
3. **Functional/drug annotation** – actionable variants, DGIdb
4. **TMB** – variant count / exome target Mb
5. **MAF analysis** – maximum allele frequency per territory

---

::: {.report-footer}
Report generated by the Somatic Variants Pipeline &nbsp;|&nbsp;
`r format(Sys.time(), "%Y-%m-%d %H:%M")` UTC  
*This report is intended for research use only. Clinical decisions should not be made solely on the basis of this automated analysis.*
:::
